---
title: "Reproducible Research: Peer Assessment 1"
author: "Jessica Miramontes"
output: 
  html_document:
  keep_md: true
---


## Loading and preprocessing the data
```{r}
stepsDS <- read.csv(file="activity.csv",head=TRUE,sep=",")
head(stepsDS)
```

## What is mean total number of steps taken per day?
First we need to make a data frame with the values needed to create a data frame with the days and the total of steps by day

```{r}
steps <- rep(NA, 61)
day <- rep("NA", 61)
stepsday <- tapply(stepsDS$steps, stepsDS$date, sum, na.rm = T)
length(stepsday)
for (i in 1:61) {
    steps[i] <- stepsday[[i]]
    day[i] <- names(stepsday)[i]
}
```
Then the dara frame is
```{r}
df <- data.frame(day, steps)
head(df)
```
The histogram is
```{r}
hist(df$steps, main = "Total steps by day", xlab = "day", col = "purple")
```

Now, the mean:
```{r}
mean(df$steps)
```

And the median:

```{r}
median(df$steps)
```



## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
time_series <- tapply(stepsDS$steps, stepsDS$interval, mean, na.rm = TRUE)
plot(row.names(time_series), time_series, type = "l", xlab = "5-min interval", 
    ylab = "Average across all Days", main = "Average number of steps taken", 
    col = "blue")

```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max_interval <- which.max(time_series)
names(max_interval)
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
activity_NA <- sum(is.na(stepsDS))
activity_NA
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

I am going to replace Na with mean in 5 min interval

```{r}
StepsAverage <- aggregate(steps ~ interval, data = stepsDS, FUN = mean)
fillNA <- numeric()
for (i in 1:nrow(stepsDS)) {
    obs <- stepsDS[i, ]
    if (is.na(obs$steps)) {
        steps <- subset(StepsAverage, interval == obs$interval)$steps
    } else {
        steps <- obs$steps
    }
    fillNA <- c(fillNA, steps)
}
```


3.Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
new_stepsDS <- stepsDS
new_stepsDS$steps <- fillNA
head(new_stepsDS)
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
StepsTotal2 <- aggregate(steps ~ date, data = new_stepsDS, sum, na.rm = TRUE)
hist(StepsTotal2$steps, main = "Total steps by day", xlab = "day", col = "pink")

```

Mean:

```{r}
mean(StepsTotal2$steps) 
mean(df$steps)


```

Median:

```{r}
median(StepsTotal2$steps) 
median(df$steps)
```

In conclusion, adding the missing values makes the median and the mean the same value which is bigger than the value without the missing numbers.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.


```{r}
day <- weekdays(as.Date(stepsDS$date))
daylevel <- vector()
for (i in 1:nrow(stepsDS)) {
    if (day[i] == "Saturday") {
        daylevel[i] <- "Weekend"
    } else if (day[i] == "Sunday") {
        daylevel[i] <- "Weekend"
    } else {
        daylevel[i] <- "Weekday"
    }
}
stepsDS$daylevel <- daylevel
stepsDS$daylevel <- factor(stepsDS$daylevel)

stepsByDay <- aggregate(steps ~ interval + daylevel, data = stepsDS, mean)
names(stepsByDay) <- c("interval", "daylevel", "steps")
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
library(lattice)
xyplot(steps ~ interval | daylevel, stepsByDay, type = "l", layout = c(1, 2), 
    xlab = "Interval", ylab = "Number of steps")

```
